---
title: "PCA Derivation with Matrix Operations"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tinytex)
#tinytex::install_tinytex(force = TRUE)  # Will use its own up-to-date packages
library(ggplot2)
library(dplyr)
```

## 1. Data setup

Let our dataset be a matrix:

\[
X \in \mathbb{R}^{n \times d}
\]

where:
- \(n\) = number of samples (rows),
- \(d\) = number of features (columns).

We first **center** the data:

\[
\tilde{X} = X - \mathbf{1}\mu^\top
\]

where \(\mu\) is the column mean vector and \(\mathbf{1}\) is an \(n\)-dimensional column of ones.

---

## 2. Covariance matrix

\[
\Sigma = \frac{1}{n} \tilde{X}^\top \tilde{X} \quad \in \mathbb{R}^{d \times d}
\]

---

## 3. Eigen decomposition

Solve:

\[
\Sigma v_i = \lambda_i v_i
\]

Collect the top \(k\) eigenvectors into the matrix:

\[
W = [v_1, v_2, \dots, v_k] \quad \in \mathbb{R}^{d \times k}
\]

- Columns of \(W\) are **orthonormal eigenvectors**.
- \(\lambda_i\) are the variances explained.

---

## 4. Projection (scores)

The **PCA scores** (projected data) are:

\[
Z = \tilde{X} W \quad \in \mathbb{R}^{n \times k}
\]

Each row of \(Z\) gives the coordinates of the data in the new \(k\)-dimensional PC space.

---

## 5. Reconstruction

We can approximately reconstruct the original data (in centered form) by projecting back:

\[
\hat{X} = Z W^\top = \tilde{X} W W^\top
\]

If \(k = d\), then \(W\) is square and orthogonal, and:

\[
\hat{X} = \tilde{X}
\]

If \(k < d\), then \(\hat{X}\) is the **rank-\(k\) approximation** of the original data.

---

## 6. Properties

1. **Orthogonality**:
   \[
   W^\top W = I_k
   \]

2. **Explained variance**:
   \[
   \text{Var}(Z_j) = \lambda_j
   \]

3. **Uncorrelated scores**:
   \[
   \text{Cov}(Z) = \Lambda_k
   \]

where \(\Lambda_k = \text{diag}(\lambda_1, \dots, \lambda_k)\).

---

```{r US arrests PCA}

mu = colMeans(USArrests)
USArrestspca = prcomp(USArrests)

nComp = 4
USArrestshat = USArrestspca$x[,1:nComp] %*% t(USArrestspca$rotation[,1:nComp])
USArrestshat = scale(USArrestshat, center = -mu, scale = FALSE)

arrests<-plot(prcomp(USArrests, scale. = TRUE))

```

