---
title: "DataPreprocessingSlidesExamples"
output: html_document
date: "2024-01-30"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
```

## Principal Component Analysis Example

Using iris data set to obtain the principal components to present the data in another space

```{r PCA Example}
# Load the iris dataset
data(iris)

# Standardize the data (optional but recommended for PCA)
scaled_data <- scale(iris[, 1:4])

# Perform PCA
pca_result <- prcomp(scaled_data)

# Summary of PCA results
summary(pca_result)

# Visualize the proportion of variance explained by each principal component
var_explained <- pca_result$sdev^2 / sum(pca_result$sdev^2)
cumulative_var <- cumsum(var_explained)

# Plot the variance explained
variance_df <- data.frame(PrincipalComponent = 1:length(var_explained), 
                          VarianceExplained = var_explained,
                          CumulativeVariance = cumulative_var)

ggplot(variance_df, aes(x = PrincipalComponent, y = VarianceExplained)) +
  geom_bar(stat = "identity", fill = "skyblue", alpha = 0.7) +
  geom_line(aes(x = PrincipalComponent, y = CumulativeVariance, group = 1), color = "red") +
  labs(title = "Variance Explained by Principal Components",
       x = "Principal Component",
       y = "Variance Explained") +
  theme_minimal()

# Visualize the data in the reduced space (first two principal components)
pca_data <- as.data.frame(pca_result$x[, 1:2])
pca_data$Species <- iris$Species

ggplot(pca_data, aes(x = PC1, y = PC2, color = Species)) +
  geom_point(size = 3) +
  labs(title = "PCA: Projection of Data onto First Two Principal Components",
       x = "Principal Component 1",
       y = "Principal Component 2") +
  theme_minimal()
```

```{r}
# Extract loadings
loadings <- pca_result$rotation

# Print the loadings
print(loadings)
```
The loadings matrix will show how each original variable contributes to each principal component. Each column of the loadings matrix corresponds to a principal component, and each row corresponds to an original variable.


To express the original variables in terms of principal components, you can use the following formula:

\[ X_{\text{original}} = Z \times \text{Loadings} \]

where:

- \( X_{\text{original}} \) is the original data matrix.
- \( Z \) is the matrix of principal components obtained from the PCA.
- \(\text{Loadings}\) is the matrix of loadings obtained from the PCA.

Here's an example using the first two principal components:

```{r}
# Extract the first two principal components
pc_scores <- pca_result$x[, 1:2]

# Express original variables in terms of principal components
original_variables <- pc_scores %*% t(loadings[, 1:2])

# Print the original variables reconstructed from the first two principal components
print(head(original_variables))
```

# Let's check mtcars data set for a quick visualization

```{r}
# Load the mtcars dataset
data(mtcars)
ggplot(data=mtcars, aes(x=hp, y=mpg)) + geom_point() + geom_smooth(method = lm)
```

Note that ggplot and *lm* might be fitting different linear models. 

```{r linear regression example}
# Load the mtcars dataset
data(mtcars)

# Fit a multiple linear regression model
model <- lm(mpg ~ hp, data = mtcars)

# Extract coefficients
coefficients <- coef(model)

# Create a scatter plot
ggplot(data = mtcars, aes(x = hp, y = mpg)) +
  geom_point() +
  geom_abline(intercept = coefficients[1], slope = coefficients[2], color = "red") +
  labs(title = "Scatter Plot with Regression Line")

```

Let's perform a multi-linear regression model to see the effect of hp and weight to mpg.
```{r multi-linear regression model}
# Load the mtcars dataset
data(mtcars)

# Fit a multiple linear regression model
model <- lm(mpg ~ hp + wt, data = mtcars)

# Extract coefficients
coefficients <- coef(model)

# Create a scatter plot
library(ggplot2)
ggplot(data = mtcars, aes(x = hp, y = mpg)) +
  geom_point() +
  geom_abline(intercept = coefficients[1], slope = coefficients[2], color = "red") +
  labs(title = "Scatter Plot with Regression Line")


```
